
---
output: 
  pdf_document:
    keep_tex: true
    fig_caption: true
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 4
title: "Project V: Parametric/Nonparametric Nonlinear Regression"
author: 
- Appiah Prince^[pappiah@miners.utep.edu]
- University of Texas at El Paso (UTEP) 
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontsize: 12pt
spacing: single
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsfonts}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{STAT 5494-- Statistical Machine Learning}
- \lhead{Parametric/Nonparametric Nonlinear Regression}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


We consider a data set jaws.txt, which is concerned about the association 
between jaw bone length (y = bone) and age in deer (x = age). We are going try 
out several parametric/nonparametric nonlinear regression models in this 
low-dimensional (p = 1) setting.

# Bring in the data and make a scatterplot of bone vs. age.

```{r}
jaw_data <- read.table(file="jaws.txt", header = TRUE)
dim(jaw_data)
head(jaw_data)
```


Scatterplot of bone vs deer
```{r}
library(ggplot2)
ggplot(data = jaw_data, mapping = aes(x = age, y = bone)) +
  geom_point()+
  geom_smooth(method = lm) +
  geom_smooth(method = "loess", col = "red") +
  ggtitle("Scatterplot of bone vs deer with linear and nonlinear fits") +
  xlab("age in deer") + ylab("jaw bone length")
```


Comment

- Almost all the points do not lie on both the linear and non linear fits.
Most of the points are also not near the curve and the line and they do not fit the data very well.

- The associations of both the linear and non linear fits do not look linear.



# Data Partitioning

## Partition the data into train and test in the ratio 2:1 respectively
```{r}
set.seed(126)
sample_jawdata <- sample(nrow(jaw_data), (2.0/3.0)*nrow(jaw_data), replace = FALSE) 
train_jawdata <- jaw_data[sample_jawdata, ] # training set
test_jawdata <- jaw_data[-sample_jawdata, ] #test set
dim(train_jawdata)
dim(test_jawdata)
```


Comment

- The training set has dimension 36 rows(observations) and 2 columns(variables).

- The test set has dimension 18 rows(observations) and 2 columns(variables).



## Checking for range of age of the train and test data to prevent extrapolation 
```{r}
range_age_original <- range(jaw_data$age)
range_age_train <- range(train_jawdata$age)
range_age_test <- range(test_jawdata$age)

range_age <- data.frame('original data'=range_age_original, 
                        'train data' =range_age_train,
                        'test data'=range_age_test)
row.names(range_age) <- c("min","max")
range_age
```


Comment

- We see that the minimum and maximum age in original data are the same as the minimum and maximum age in the training set. 

- We also observe that the range of age in the test set does not exceed that in the training set.

- Hence,the problem of extrapolation when it comes to prediction is prevented.



# Parametric Nonlinear Models

## Fitting an asymptotic exponential model
```{r}
bone_jaw_model <- nls(bone ~ beta1 - beta2*exp(-beta3*age),data=train_jawdata,
    start=list(beta1 =120, beta2 =5, beta3 = 0.3), trace=T)
summary(bone_jaw_model)
```


Comment

All the coefficients are statistically significant according to their p-values at 5% significant level. Also, the number of iterations to convergence is 6 with the Achieved convergence tolerance of 8.58e-06.


## Fitting the reduced model under H0.
```{r}
reduced_bone_jaw_model <- nls(bone ~ beta1*(1-exp(-beta3*age)),
            data=train_jawdata,start=list(beta1 = 120, beta3 = 0.3), trace=T)
summary(reduced_bone_jaw_model)
```


Comment

All the coefficients are statistically significant according to their p-values 
at 5% significant level. Also, the number of iterations to convergence is 4 
with the Achieved convergence tolerance of 8.394e-06.


```{r}
anova(bone_jaw_model, reduced_bone_jaw_model)
```


Comment

The p-value 0.7583 is greater than 0.05 which implies that we fail to reject
H0.


```{r}
AIC1 <- AIC(bone_jaw_model)
BIC1 <- BIC(bone_jaw_model)
AIC2 <- AIC(reduced_bone_jaw_model)
BIC2 <- BIC(reduced_bone_jaw_model)

AICS <- c(AIC1,AIC2,BIC1,BIC2)
compare <- data.frame(Methods=c('AIC non-reduced model','AIC reduced model',
                                'BIC non-reduced model','BIC reduced model'),
                                'Values' =AICS )
knitr::kable(compare, align = "lc", caption = "Comparing the two nls models 
             with AIC and BIC ")
```



Comment

Both the AIC and BIC of the reduced model are smaller than that of the 
non-reduced models'. Thus, this result together with the result from the anova 
function and the reduced model's parsimony confirm that the reduced model is 
better that the non-reduced model.


## Based on the better model, add the  fitted curve to the scatterplot.

```{r}
fitted_values <- fitted.values(reduced_bone_jaw_model)
library(ggplot2)
ggplot(data = train_jawdata, mapping = aes(x = age, y = bone)) +
  geom_point()+
  geom_line(aes(x=age, y=fitted_values), col = 'red') +
  ggtitle("fitted curve from our best model on the scatterplot") +
  xlab("age in deer") + ylab("jaw bone length")
```


Comment

The fitted curve to some extent fits the data well and also few points are far 
away from the curve.


Apply the better model to the test set.
```{r}
y_hat <- predict(reduced_bone_jaw_model, newdata = test_jawdata)
```


Plot the observed y values in test set versus their predicted values together
with the reference line y = x, to check if the prediction seems reasonable.

```{r}
y_test_jawdata <- test_jawdata$bone
ggplot( mapping = aes(x =y_test_jawdata , y = y_hat)) +
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")

```


Comment

Most of the points are far away from the reference line which indicates that the predicted values are significantly different from the observed values in the 
test data.


Computing the prediction mean square error (MSE)
```{r}
prediction_MSE <- mean((y_test_jawdata - y_hat)^2)
prediction_MSE
```
The prediction mean square error (MSE) is 175.4818


\newpage
# Local regression methods

## KNN regression model

V-FOLD CV FOR SELECTING K
```{r}
set.seed(123)
library("FNN")
SSEP <- function(yobs, yhat) sum((yobs-yhat)^2) 

K <- 2:15
V <- 6
# id.fold <- sample(1:V, size = NROW(train_jawdata), replace=T)
SSE <- rep(0, length(K))
for(k in 1:length(K)){
  id.fold <- sample(rep(1:V, each=trunc(NROW(train_jawdata)/V)))
  for(v in 1:V){
    train1<- train_jawdata[id.fold!=v, ];
    train2<- train_jawdata[id.fold==v, ];
    yhat2 <- knn.reg(train=train1, y=train1$bone, test=train2, k=K[k], algorithm="kd_tree")$pred;
    SSE[k] <- (SSE[k] + SSEP(train2$bone, yhat2))
  }  
}
cbind(K, SSE)
k.opt <- K[which.min(SSE)]
k.opt
```


Comment

To avoid over-fitting(that is when K=1), I choose K to range from 2 to 15 and 
using the 6-fold CV, we see that K=2, gives the optimal K. 

```{r}
knn_reg_model <- knn.reg(train=train_jawdata, y=train_jawdata$bone,
                         k=k.opt, algorithm="kd_tree")
summary(knn_reg_model)
#names(knn_reg_model)
```

Plot the fitted curve together with the scatterplot of the data.
```{r}
library(ggplot2)
ggplot(data = train_jawdata, mapping = aes(x = age, y = bone)) +
  geom_point()+
  geom_line(aes(x=age, y=knn_reg_model$pred), col = 'blue') +
  ggtitle("fitted curve from knn_reg_model on the scatterplot") +
  xlab("age in deer") + ylab("jaw bone length")
```


Comment

The plot seems to fit the data very well. As wee can see the curve almost 
wiggles through all the data points.


```{r}
knn_reg_model2 <- knn.reg(train=train_jawdata, test = test_jawdata, 
                          y=train_jawdata$bone, k=k.opt, algorithm="kd_tree")
```



Plot the observed and predicted response values with reference line y = x
```{r}
y_test_jawdata <- test_jawdata$bone
library(ggplot2)
ggplot( mapping = aes(x =y_test_jawdata , y = knn_reg_model2$pred)) +
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")

```


Comment

We observe that the points are near to the reference line which indicates that 
the observed values in thee test data is significant not different from the 
predicted values. This indicates a good prediction.


Computing the prediction mean square error (MSE)
```{r}
prediction_MSE1 <-mean((y_test_jawdata - knn_reg_model2$pred)^2)
prediction_MSE1
```

Comment

The prediction mean square error (MSE) is 12.59012 which is small which 
indicates the error of prediction is minimized.



## Applying kernel regression to obtain a nonlinear fit.

```{r}
library(lokern)
lofit <- lokerns(train_jawdata$age, train_jawdata$bone)
sb <- summary(lofit$bandwidth)

op <- par(fg = "gray90", tcl = -0.2, mgp = c(3,.5,0))
plot(lofit$band, ylim=c(0,3*sb["Max."]), type="h", ann = F, axes = FALSE)
#if(R.version$major > 1 || R.version$minor >= 3.0)
boxplot(lofit$bandwidth, add = TRUE, at = 304, boxwex = 8,
    col = "gray90", border="gray", pars = list(axes = FALSE))
    axis(4, at = c(0,pretty(sb)), col.axis = "gray")
par(op) 
par(new=TRUE)

plot(bone ~ age, data = train_jawdata, main = "Local Plug-In Bandwidth Vector")
lines(lofit$x.out, lofit$est, col=3)
mtext(paste("bandwidth in [", paste(format(sb[c(1,6)], dig = 3),collapse=","),
    "];  Median b.w.=",formatC(sb["Median"])))
```


Comment

- We used Kernel Regression Smoothing with Local Plug-in Bandwidth.

- The bandwidth is within the interval [4.31,8.24] with median bandwidth of 
6.495.


Apply the fitted kernel regression model to the test data.
```{r}
perdict_kernel_reg <- predict(lofit, newdata = test_jawdata)
```


Plot the observed and predicted response values with reference line y = x
```{r}
library(ggplot2)
ggplot( mapping = aes(x =y_test_jawdata , y = perdict_kernel_reg$y)) +
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")

```


Comment

We see from the plot that most of the points are far from the reference line 
which indicates poor prediction. That is the predicted values are statistically
different from the observed values in the test data.


Computing the prediction MSE
```{r}
prediction_MSE2 <- mean((y_test_jawdata - perdict_kernel_reg$y)^2)
prediction_MSE2
```

Comment

The prediction MSE is 670.7431 which is large.



## local (cubic) polynomial regression

```{r, warning=FALSE}
library(locpol)
fit.local <- locpol(bone~age, data=train_jawdata, deg=3, kernel=EpaK,bw =4)
```

Comment

- The kernel used here is EpaK

- The bandwidth used is 4


Apply the local cubic regression model to the test data.
```{r, warning=FALSE}
lp <- locpol(bone~age, data=train_jawdata, xeval=test_jawdata$age, deg=3,
             kernel=EpaK,bw =7)
lpp <- lp$lpFit
perdict_locpol <- lpp$bone
```

Comment

Here, I used bandwidth of 7 since I couldnt run with a bandwidth of 4.


Plot the observed and predicted response values with reference line y = x
```{r}
library(ggplot2)
ggplot( mapping = aes(x =y_test_jawdata , y = perdict_locpol)) +
  geom_point()+
  geom_abline(intercept = 0, slope = 1, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")
```


Comment

We see from the plot that most of the points are far from the reference line 
which indicates poor prediction. That is the predicted values are statistically
different from the observed values in the test data.


Computing Prediction MSE.
```{r}
y_test_jawdata <- test_jawdata$bone
prediction_MSE3 <- mean((y_test_jawdata - perdict_locpol)^2)
prediction_MSE3
```


Comment

The prediction MSE is 283.0541 which is relatively large.


# Regression/smoothing splines


## Regression splines(Natural cubic splines)

```{r}
library(splines)
#ns(train_jawdata$age, df = 5)
natural_cubspl <- lm(bone ~ ns(age, df = 5), data = train_jawdata)
```


Plot the resultant curve.
```{r}
library(ggplot2)
ggplot(data = train_jawdata, mapping = aes(x = age, y = bone)) +
  geom_point()+
  geom_line(aes(x=age, y=natural_cubspl$fitted.values), col = 'red') +
  ggtitle("fitted curve from natural cubic spline model on the scatterplot") +
  xlab("age in deer") + ylab("jaw bone length")
```


Comment

The curve does not fit the data very well. Most points are not near the curve.


Applying the fitted model to predict the test data.
```{r}
y_hatt <- predict(natural_cubspl, newdata = test_jawdata)
```


Plot the observed and predicted response values with reference line y=x.
```{r}
library(ggplot2)
ggplot( mapping = aes(x =test_jawdata$bone , y = y_hatt)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")
```


Comment

We see from the plot that most of the points are far from the reference line
which indicates poor prediction. That is the predicted values are statistically
different from the observed values in the test data.


Computing the prediction MSE
```{r}
prediction_MSE4 <- mean((test_jawdata$bone - y_hatt)^2)
prediction_MSE4
```

Comment

The prediction MSE is 200.7106 which relatively large


## Smoothing splines
```{r}
library(splines)
smooth_spline <- smooth.spline(train_jawdata$age, train_jawdata$bone)
smooth_spline
```


Comment

The generalized cross-validation (GCV) was used for the smoothing 
parameter estimation. We obtained GCV of 3892.639 and an Equivalent Degrees 
of Freedom (Df) of 5.508976.



Add the resultant curve to the scatterplot.
```{r}
library(ggplot2)
ggplot(data = train_jawdata, mapping = aes(x = age, y = bone)) +
  geom_point()+
  geom_line(aes(x=age, y=fitted(smooth_spline)), col = 'red') +
  ggtitle("fitted curve from smoothing spline model on the scatterplot") +
  xlab("age in deer") + ylab("jaw bone length")
```


Comment

The curve does not fit the data very well. Most points are not near the curve.


Apply the fitted model to the test data.
```{r}
y_hat_sp <- predict(smooth_spline, test_jawdata$age)
```


Plot the observed and predicted response values
```{r}
library(ggplot2)
ggplot( mapping = aes(x =test_jawdata$bone , y = y_hat_sp$y)) +
  geom_point()+
  geom_abline(slope = 1, intercept = 0, col = 'red')+
  ggtitle("Plot of observed y in test data vs predicted y") +
  xlab("observed y in test data") + ylab("predicted y")

```


Comment

From the bottom to the middle part of the reference line we see that the points 
are not near the line. In particular the predicted values are statistically 
different from the observed values in the test data.


Compute the Prediction MSE
```{r}
prediction_MSE5 <- mean((test_jawdata$bone - y_hat_sp$y)^2)
prediction_MSE5
```


Comment

The prediction MSE is 177.413 which is relatively large.



# Tabulate all the prediction MSE measures


```{r}
measure <- c(prediction_MSE, prediction_MSE1, 
             prediction_MSE2,prediction_MSE3,
             prediction_MSE4,prediction_MSE5)
method_measure <- data.frame("Methods"= c("Asymptotic exponential model",
                              "KNN regression","Kernel regression",
                              "Local cubic polynomial","Natural cubic spline",
                              "Smoothing Splines"), "Prediction_MSE"= measure)
knitr::kable(method_measure, align = "lc", 
             caption = "A table of all the prediction MSE measures.")
```

From the output of the prediction MSE for various methods, we see that KNN regression has the smallest  prediction MSE. Hence, KNN regression gives the favorable result. 